Review of Edge Detectors

### 2014-11-11 old_versions/edge-detection-review-v5.tex
second round of third revision, improving references

%-------------------------------------------------------------------------------
% Edge detectors review
% by Haldo Spontón & Juan Cardelino
% IPOL 2012
%-------------------------------------------------------------------------------
\documentclass{ipol}
\ipolSetYear{2012}
\ipolSetDOI{10.XXXX/ipol.2012.sc-red}

\usepackage{hyperref,verbatim,graphicx,amsmath,amssymb,amssymb,dsfont}
\usepackage[ruled,linesnumbered]{algorithm2e}
\usepackage{sidecap,subfig,array}
\usepackage[table]{xcolor}
\usepackage{multicol}
\usepackage{listings}
\usepackage{color}
\newtheorem{theorem}{Theorem}
\numberwithin{equation}{section}
\numberwithin{table}{section}
\numberwithin{figure}{section}
\begin{document}
\bibliographystyle{ieeetr}
\lstset{
	backgroundcolor=\color[rgb]{0.9,0.9,0.9},
    tabsize=4,
    inputencoding=utf8x,
         extendedchars=\true,
         language=C++,
         basicstyle=\scriptsize,
                 showtabs=false,
        showspaces=false,
        showstringspaces=false,
        identifierstyle=\ttfamily,
        keywordstyle=\color[rgb]{0,0,1},
        commentstyle=\color[rgb]{0.133,0.545,0.133},
        stringstyle=\color[rgb]{0.627,0.126,0.941},
}

@@

%-------------------------------------------------------------------------------
\begin{ipolAbstract}
In this paper we study some of the classic alternatives for edge detection in digital images. The main idea 
behind the edge detection is to find where there are abrupt changes in the intensity of an image. 
As a first approach, the first derivative is used to find the changes of intensity in several algorithms 
such as Sobel, Prewitt and Roberts \cite{Gonzalez2007Digital}. The second derivative is also used, for example in algorithms 
like Marr-Hildreth \cite{AIM-518} and Haralick \cite{bb20239}.\\
We compare the results obtained from a qualitative point of view (perceptual) and from a quantitative 
point of view (number of operations, execution time), considering different forms to convolve an 
image with a kernel (step required in some of the algorithms).\\

%The Marr-Hildreth algorithm is a method of detecting edges in digital 
%images. It's based on convolving the image with the Laplacian of the 
%Gaussian function. Then, zero crossings are detected in the filtered 
%result to obtain the edges points.
\end{ipolAbstract}

%-------------------------------------------------------------------------------
\begin{ipolCode}
The C implementation of the algorithms implemented in this work (beta version) can be
downloaded from \href{http://iie.fing.edu.uy/~haldos/downloads/edge_detectors_v0.1.tar.gz}{here}.
%\begin{itemize}
%	\item Roberts, Prewitt and Sobel operators: \href{http://iie.fing.edu.uy/~haldos/ipol/fded.tar.gz}{here}.
%	\item Marr-Hildreth algorithm: \href{http://iie.fing.edu.uy/~haldos/ipol/marr-hildreth.tar.gz}{here}.
%	\item Haralick algorithm: \href{http://iie.fing.edu.uy/~haldos/ipol/haralick.tar.gz}{here}.
%\end{itemize}
% TODO referencia a github
\end{ipolCode}

@@

In the 70's, edge detection methods were based on using small operators 
(such as Sobel masks), attempting to compute an approximation of the
first derivative of the image. In section ~\ref{sec:first} we study the 
performance of such algorithms, as an introduction to a more sophisticated 
analysis into the edge finding process.\\

In 1980 Marr and Hildreth argued that intensity changes are not independent 
of image scale, so the edge detection requires the use of different size 
operators. They also argued that a sudden intensity change will be seen 
as a peak (or trough) in the first derivative or, equivalently, as a zero 
crossing in the second derivative. The bulk of this work focuses on the 
study of this algorithm, presented in Section ~\ref{sec:second}. In this section 
we also present the Haralick algorithm, which is also based on the second derivative.
This algorithm has the particularity of proposing a model to locally approximate 
the image around a point. Then, with this model, second derivative can be calculated 
analytically (rather approximate) and finding edges is done by imposing a 
condition over the model parameters. \\

Results and conclusions are shown in sections \ref{sec:results} and 
\ref{sec:conclusions} respectively. In section ~\ref{sec:examples} we present some examples to compare the 
performance of the algorithms studied in this work. We also present a 
video created using the implemented algorithms, applied frame by frame.\\

%-------------------------------------------------------------------------------
\section{Algorithms based on first derivative}
\label{sec:first}

@@

The two algorithms that we will see in this section (Marr-Hildreth [1980] and Haralick [1987]) 
are based on the second derivative of the image, and both take steps to reduce noise before 
detecting edges in the image.\\

@@

The Marr-Hildreth algorithm is a method of detecting edges in digital 
images. It's based on find zero crossing points of the second derivative
of the image. This can be done in several ways. In this work we implemented 
two different ways; convolving the image with a Gaussian kernel and then 
approximating the second derivative (Laplacian) with a 3x3 kernel, or 
convolving the image with a kernel calculated as the Laplacian of a 
Gaussian function. There are more ways to do so, for example, using 
recursive Gaussian filters.\\ % poner referencia!

@@

The idea behind the Haralick edge detector is identical to that of the previous method; find zeros in 
the second derivative of the image. In this method, we smooth the input image through local bi-cubic
polynomial fitting. Then, calculating the second derivative analytically, we can find an expression 
equivalent to cancel the second derivative of the polynomial as a function of the parameters of it.\footnote{This implementation is slightly 
different to the traditional implementation of the Haralick algorithm. We do not use the condition concerning the third derivative.}\\

@@

First we see the results of each of the algorithms on a simple test image, the white square on black background shown in figure \ref{fig:original1}. \\

@@

	\subfloat[Marr-Hildreth (LoG).]{\label{fig:result2-e}\includegraphics[width=0.42\textwidth]{results/molino_marr-hildreth-log.png}}
	\quad
	\subfloat[Haralick.]{\label{fig:result2-f}\includegraphics[width=0.42\textwidth]{results/molino_haralick.png}}
	\caption{Results of the first derivative, Marr-Hildreth and Haralick algorithms using the figure \ref{fig:original2} as input image.}
	\label{fig:result2}
\end{figure}

\begin{table}[h!]
	\begin{center}
	\begin{tabular}{| l | r |}
		\hline \rule{0pt}{3ex}
		\cellcolor[gray]{0.8} \textbf{Algorithm}	& \cellcolor[gray]{0.8} \textbf{Execution time (s)}	\\ \hline \rule{0pt}{3ex}
		Roberts, Prewitt and Sobel					& $0.550 \ s$										\\ \hline \rule{0pt}{3ex}
		Marr-Hildreth (Gaussian)					& $1.050 \ s$										\\ \hline \rule{0pt}{3ex}
		Marr-Hildreth (LoG)							& $1.440 \ s$										\\ \hline \rule{0pt}{3ex}
		Haralick									& $0.930 \ s$										\\
		\hline
	\end{tabular}
	\end{center}
	\caption{Execution time of the algorithms using the figure \ref{fig:original2} as input image (1000$\times$563 pixels, 3 channels).}
	\label{exectime2}
\end{table}
\vspace{0.5cm}

It is now visible the difference in performance between the first derivative edge detection algorithms and the second derivative ones. We can see in figure \ref{fig:result3} an area of ​​interest of the image, enlarged to better view of the detail. Note that none of the first derivative methods (even with a loose threshold) detect the lower edge of the blade (which is shaded). Neither Haralick algorithm detects it. However, the Marr-Hildreth algorithm detects it, using either a Gaussian or a LoG kernel. \\

Another observation is that the edges detected by the Haralick algorithm appear to be thicker than those detected with the other algorithms. This may be due the regularity that Haralick assumes, previously mentioned. \\

@@

	\subfloat[Marr-Hildreth (LoG).]{\label{fig:result3-f}\includegraphics[width=0.48\textwidth]{results/molino_marr-hildreth-log_zoom.png}}
	\quad
	\subfloat[Haralick.]{\label{fig:result3-g}\includegraphics[width=0.48\textwidth]{results/molino_haralick_zoom.png}}
	\caption{Results of the algorithms using the figure \ref{fig:original2} as input image, enlarging an interesting area.}
	\label{fig:result3}
\end{figure}

@@

We have discussed and carefully implemented some of the more traditional methods of edge detection in digital images. We have tested the implemented algorithms with synthetic and real images, obtaining generally the expected results, taking into account the limitations of these methods. \\

@@

Comparing the two versions of the Marr-Hildreth algorithm, is significantly faster the version with Gaussian kernel and numerical approximation of the Laplacian. The LoG kernel version, while slower (since it needs a larger kernel to achieve similar results) is more accurate, because it makes no approximation to calculate the Laplacian (it is calculated analytically before creating the kernel). It is also possible to manage the size of the kernel, which represents a scale parameter of the algorithm, being able to obtain a highly detailed edge image using small kernels, or just more noticeable edges using larger kernels. \\

@@

All algorithms have their advantages and disadvantages. Choosing one or other may depend on requirements after edge detection. \\

% At the end:
Finally, this paper is a summary of some classical edge detection algorithms. In the preparation of this paper, no similar material (concentrated in one text) was found, so this paper appears to be useful material for whom starts studying edge detection methods. \\

@@

Some observations: 
\begin{itemize}
	\item The first derivative algorithms, although they work properly and quickly, have some problems like discontinuity in the edges. They are also very affected by noise in images, appearing unwanted isolated edges. \\
	\item The Marr-Hildreth algorithm (in its two versions) achieved interesting results in the edge detail. This algorithm provides a first step to regularity in the image, since the filtering with a Gaussian kernel. This behavior can be seen in Lena's hair (figure \ref{fig:example1}), or in the inside of the oranges (example \ref{fig:example2}). \\
	\item As mentioned earlier, Haralick's algorithm is the first to assume greater regularity of the image in the neighborhood of a pixel (third order). This causes thicker edges, as shown in both examples. But, note that some edges are only detected with this algorithm, e.g. the lower edges of the oranges, which are in the shadow, in figure \ref{fig:example2}. These edges are quite smooth, so that the Haralick model is well suited to them, and they do not represent an abrupt change in the intensity to be detected by other methods. \\
\end{itemize}

%\vspace{2cm}

%\begin{figure}[t!]
%	\centering
%	\subfloat[Original.]{\includegraphics[width=0.3\textwidth]{examples/shapes.png}}

%	\subfloat[Roberts.]{\includegraphics[width=0.3\textwidth]{examples/shapes_roberts.png}}
%	\quad
%	\subfloat[Prewitt.]{\includegraphics[width=0.3\textwidth]{examples/shapes_prewitt.png}}
%	\quad
%	\subfloat[Sobel.]{\includegraphics[width=0.3\textwidth]{examples/shapes_sobel.png}}

%	\subfloat[Marr-Hildreth (Gaussian).]{\includegraphics[width=0.3\textwidth]{examples/shapes_marr-hildreth.png}}
%	\quad
%	\subfloat[Marr-Hildreth (LoG).]{\includegraphics[width=0.3\textwidth]{examples/shapes_marr-hildreth-log.png}}
%	\quad
%	\subfloat[Haralick.]{\includegraphics[width=0.3\textwidth]{examples/shapes_haralick.png}}
%	\caption{Example: Shapes.}
%	\label{fig:example3}
%\end{figure}

@@

%TODO: videos de ejemplo con todos los métodos.

This is an example of applying the edge detection algorithms described here, frame by frame, to a video: 
\begin{itemize}
	\centering
	\item \href{http://iie.fing.edu.uy/~haldos/ipol/video.mov}{original} (xxxMb).
\end{itemize}
\begin{multicols}{2}
\begin{itemize}
	\item \href{http://iie.fing.edu.uy/~haldos/ipol/video-roberts.mov}{roberts version} (xxxMb).
	\item \href{http://iie.fing.edu.uy/~haldos/ipol/video-prewitt.mov}{prewitt version} (xxxMb).
	\item \href{http://iie.fing.edu.uy/~haldos/ipol/video-sobel.mov}{sobel version} (xxxMb).
	\item \href{http://iie.fing.edu.uy/~haldos/ipol/video-marr-hildreth-gaussian.mov}{marr-hildreth-gaussian version} (xxxMb).
	\item \href{http://iie.fing.edu.uy/~haldos/ipol/video-marr-hildreth-log.mov}{marr-hildreth-log version} (xxxMb).
	\item \href{http://iie.fing.edu.uy/~haldos/ipol/video-haralick.mov}{haralick version} (xxxMb).
\end{itemize}
\end{multicols}

@@

%-------------------------------------------------------------------------------
\subsection{Kernel generation}
\label{sec:appendix2}

@@

>>>>

%-------------------------------------------------------------------------------
% Edge detectors review v5
% by Haldo Spontón & Juan Cardelino
% IPOL 2012
%-------------------------------------------------------------------------------
\documentclass{ipol}
\ipolSetYear{2012}
\ipolSetDOI{10.XXXX/ipol.2012.sc-red}

\usepackage{hyperref,verbatim,graphicx,amsmath,amssymb,amssymb,dsfont}
\usepackage[ruled,linesnumbered]{algorithm2e}
\usepackage{sidecap,subfig,array}
\usepackage[table]{xcolor}
\usepackage{multicol}
\usepackage{listings}
\usepackage{color}
\newtheorem{theorem}{Theorem}
\numberwithin{equation}{section}
\numberwithin{table}{section}
\numberwithin{figure}{section}
\begin{document}
\bibliographystyle{ieeetr}
\lstset{
	backgroundcolor=\color[rgb]{0.94,0.94,0.94},
    tabsize=4,
    inputencoding=utf8x,
         extendedchars=\true,
         language=C,
         basicstyle=\scriptsize,
                 showtabs=false,
        showspaces=false,
        showstringspaces=false,
        identifierstyle=\ttfamily,
        keywordstyle=\color[rgb]{0,0,1},
        commentstyle=\color[rgb]{0.133,0.545,0.133},
        stringstyle=\color[rgb]{0.627,0.126,0.941},
}

@@

%-------------------------------------------------------------------------------
\begin{ipolAbstract}
In this paper we study some of the classic alternatives for edge detection in digital images. The main idea 
behind edge detection is to find where there are abrupt changes in the intensity of an image. 
As a first approach, the first derivative is used to find the changes of intensity in several algorithms 
such as Sobel, Prewitt and Roberts. The second derivative is also used, for example in algorithms 
like Marr-Hildreth and Haralick.\\
We compare the results obtained from a qualitative point of view (perceptual) and from a quantitative 
point of view (number of operations, execution time), considering different forms to convolve an 
image with a kernel (step required in some of the algorithms).\\
\end{ipolAbstract}

%-------------------------------------------------------------------------------
\begin{ipolCode}
The C implementation of the algorithms implemented in this work (beta version) can be
downloaded from \href{http://iie.fing.edu.uy/~haldos/downloads/edge_detectors_v0.1.tar.gz}{here}.
You can also follow this code project in this GIT repository: 
\begin{center}
\url{git://github.com/haldos/edges.git}
\end{center}
\end{ipolCode}

@@

In the 70's, edge detection methods were based on using small operators 
(such as Sobel masks), attempting to compute an approximation of the
first derivative of the image \cite{Gonzalez2007Digital}. In section ~\ref{sec:first} we study the 
performance of such algorithms, as an introduction to a more sophisticated 
analysis into the edge finding process.\\

In 1980 Marr and Hildreth \cite{AIM-518} argued that intensity changes are not independent 
of image scale, so the edge detection requires the use of different size 
operators. They also argued that a sudden intensity change will be seen 
as a peak (or trough) in the first derivative or, equivalently, as a zero 
crossing in the second derivative. The bulk of this work focuses on the 
study of this algorithm, presented in Section ~\ref{sec:second}. In this section 
we also present the Haralick algorithm \cite{bb20239}, which is also based on the second derivative.
This algorithm has the particularity of proposing a model to locally approximate 
the image around a point. Then, with this model, the second derivative can be calculated 
analytically (rather approximate) and finding edges is done by imposing a 
condition over the model parameters. \\

Results and conclusions are shown in sections \ref{sec:results} and 
\ref{sec:conclusions} respectively. In section ~\ref{sec:examples} we present some examples to compare the 
performance of the algorithms studied in this work. \\

In conjunction with this paper we present a detailed and well commented source code, which 
implements the described algorithms. Also in the appendices (sections \ref{sec:appendix1} 
and \ref{sec:appendix2}) we show some common mathematical developments, and a commented 
explanation of the source code. This work also provides tools for code documentation 
(see section \ref{sec:appendix2}). \\

\nocite{IPOL}
%-------------------------------------------------------------------------------
\section{Algorithms based on first derivative}
\label{sec:first}

@@

The two algorithms that we will see in this section (Marr-Hildreth \cite{AIM-518} [1980] and Haralick \cite{bb20239} [1987]) 
are based on the second derivative of the image, and both take steps to reduce noise before 
detecting edges in the image.\\

@@

The Marr-Hildreth algorithm is a method of detecting edges in digital 
images. It's based on find zero crossing points of the second derivative
of the image. This can be done in several ways. In this work we implemented 
two different ways; convolving the image with a Gaussian kernel and then 
approximating the second derivative (Laplacian) with a 3x3 kernel, or 
convolving the image with a kernel calculated as the Laplacian of a 
Gaussian function. There are more ways to do so, for example, using 
recursive Gaussian filters \cite{Deriche1993Recursively}.\\

@@

The idea behind the Haralick edge detector is identical to that of the previous method; find zeros in 
the second derivative of the image. In this method, we smooth the input image through local bi-cubic
polynomial fitting. Then, calculating the second derivative analytically, we can find an expression 
equivalent to cancel the second derivative of the polynomial as a function of the parameters of it.\footnote{This implementation is slightly 
different to the traditional implementation of the Haralick algorithm. We do not use the condition concerning the third derivative (to be negative), in order to implement two edge detection methods using only the second derivative (along the Marr-Hildreth algorithm).}\\

@@

First we show the results of each algorithm on a simple test image, the white square on black background shown in figure \ref{fig:original1}. \\

@@

	\subfloat[Marr-Hildreth (LoG).]{\label{fig:result2-e}\includegraphics[width=0.42\textwidth]{results/molino_marr-hildreth-log.png}}
	\quad
	\subfloat[Haralick.]{\label{fig:result2-f}\includegraphics[width=0.42\textwidth]{results/molino_haralick.png}}
	\caption{Results of the first derivative, Marr-Hildreth and Haralick algorithms using figure \ref{fig:original2} as input image.}
	\label{fig:result2}
\end{figure}

\begin{table}[h!]
	\begin{center}
	\begin{tabular}{| l | r |}
		\hline \rule{0pt}{3ex}
		\cellcolor[gray]{0.8} \textbf{Algorithm}	& \cellcolor[gray]{0.8} \textbf{Execution time (s)}	\\ \hline \rule{0pt}{3ex}
		Roberts, Prewitt and Sobel					& $0.550 \ s$										\\ \hline \rule{0pt}{3ex}
		Marr-Hildreth (Gaussian)					& $1.050 \ s$										\\ \hline \rule{0pt}{3ex}
		Marr-Hildreth (LoG)							& $1.440 \ s$										\\ \hline \rule{0pt}{3ex}
		Haralick									& $0.930 \ s$										\\
		\hline
	\end{tabular}
	\end{center}
	\caption{Execution time of the algorithms using figure \ref{fig:original2} as input image (1000$\times$563 pixels, 3 channels).}
	\label{exectime2}
\end{table}

In this example the difference in performance between the first derivative edge detection algorithms and the second derivative ones is more significative. We can see in figure \ref{fig:result3} an area of ​​interest of the image, enlarged to have a better view of the detail. Note that none of the first derivative methods (even with a loose threshold) detect the lower edge of the blade (which is shaded). Neither Haralick algorithm detects it. However, the Marr-Hildreth algorithm detects it, using either a Gaussian or a LoG kernel. \\

Another observation is that edges detected by Haralick algorithm appear to be thicker than those detected with the other algorithms. This may be due the regularity that Haralick assumes, we mentioned 
before. \\

@@

	\subfloat[Marr-Hildreth (LoG).]{\label{fig:result3-f}\includegraphics[width=0.48\textwidth]{results/molino_marr-hildreth-log_zoom.png}}
	\quad
	\subfloat[Haralick.]{\label{fig:result3-g}\includegraphics[width=0.48\textwidth]{results/molino_haralick_zoom.png}}
	\caption{Results of the algorithms using figure \ref{fig:original2} as input image, enlarging an interesting area.}
	\label{fig:result3}
\end{figure}

@@

We have discussed and carefully implemented some of the most traditional methods of edge detection in digital images. We have tested the implemented algorithms with synthetic and real images, obtaining generally the expected results, taking into account the limitations of these methods. \\

@@

Comparing the two versions of the Marr-Hildreth algorithm, the version with Gaussian kernel runs significative faster that the LoG one. The latter, while slower (since it needs a larger kernel to achieve similar results) is more accurate, because it makes no approximation to calculate the Laplacian (it is calculated analytically before creating the kernel). It is also possible to manage the size of the kernel, which represents a scale parameter of the algorithm, being able to obtain a highly detailed edge image using small kernels, or just more noticeable edges using larger kernels. \\

@@

All algorithms have their advantages and disadvantages. Choosing one or other may depend on requirements of the application. \\

To conclude, this paper is a summary of some classical edge detection algorithms. In the preparation of this paper, no similar material (concentrated in one text) was found, so this paper appears to be useful material for whom starts studying edge detection methods. \\

As an addendum to this work we present a detailed implementation of the described algorithms, 
available in C language, along with an online demo where users can run the algorithms 
with arbitrary images uploaded by themselves. We also supply some code documentation tools, 
which were used to generate the detailed code description presented in section \ref{sec:appendix2} (Appendix 2). We think this tool could be useful for future IPOL articles. \\

@@

Some observations: 
\begin{itemize}
	\item The first derivative algorithms, although they work properly and quickly, have some problems like discontinuity in the edges. They are also very affected by noise in images, appearing unwanted isolated edges (More sophisticated methods help to avoid and improve this, e.g. Canny edge detector \cite{Canny1986}). \\
	\item The Marr-Hildreth algorithm (in its two versions) achieved interesting results in the edge detail. This algorithm provides a first step to regularity in the image, since the filtering with a Gaussian kernel. This behavior can be seen in Lena's hair (figure \ref{fig:example1}), or inside the oranges (example \ref{fig:example2}). \\
	\item As mentioned earlier, Haralick's algorithm is the first to assume greater regularity of the image in the neighborhood of a pixel (third order). This causes thicker edges, as shown in both examples. But, note that some edges are only detected with this algorithm, e.g. the lower edges of the oranges, which are in the shadow, in figure \ref{fig:example2}. These edges are quite smooth, so that the Haralick model is well suited to them, and they do not represent an abrupt change in the intensity to be detected by other methods. \\
\end{itemize}

@@

This is an example of applying the edge detection algorithms described here, frame by frame, to a video: 
\begin{itemize}
	\centering
	\item \href{http://iie.fing.edu.uy/~haldos/ipol/video.mov}{original} (43 Mb).
\end{itemize}
\begin{multicols}{2}
\begin{itemize}
	\item \href{http://iie.fing.edu.uy/~haldos/ipol/video-roberts-wide_0.1.mov}{roberts version} (7.9 Mb).
	\item \href{http://iie.fing.edu.uy/~haldos/ipol/video-prewitt-wide_0.1.mov}{prewitt version} (9.1 Mb).
	\item \href{http://iie.fing.edu.uy/~haldos/ipol/video-sobel-wide_0.1.mov}{sobel version} (9.1 Mb).
	\item \href{http://iie.fing.edu.uy/~haldos/ipol/video-marr-hildreth-gaussian-wide_3_19_0.04.mov}{marr-hildreth-gaussian version} (12 Mb).
	\item \href{http://iie.fing.edu.uy/~haldos/ipol/video-marr-hildreth-log-wide_3_25_0.04.mov}{marr-hildreth-log version} (14 Mb).
	\item \href{http://iie.fing.edu.uy/~haldos/ipol/video-haralick-wide_0.5.mov}{haralick version} (12 Mb).
\end{itemize}
\end{multicols}

@@

%-------------------------------------------------------------------------------
\subsection{Kernel generation}

@@

This code documentation was generated using a perl script to extract some parts of the 
code and comments into latex files. This code documentation tool is available as a GIT
project: 
\begin{center} 
\url{https://github.com/juan-cardelino/source_comment_extractor}. \\
\end{center}
Also Doxygen code documentation is available:
\begin{center}
\url{http://iie.fing.edu.uy/~haldos/ipol/red_v0.1}
\end{center}
